Visualizing Contingency Tables

### 2006-05-03 trunk/vcd/paper/hsc.Rnw
re-arranged HSC graphics, handled dimnames of tables consistently
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@529 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

<<preliminaries,echo=FALSE,results=hide>>=
myseed <- 1071
set.seed(myseed)
library(vcd)
library(scatterplot3d)
library(xtable)
data(Arthritis)
data(UCBAdmissions)
hospital <- as.table(matrix(c(43,16,3,6,11,10,9,18,16), c(3,3), byrow = TRUE, 
dimnames = list("Visit frequency" = c("Regular","Less than monthly","Never"), 
"Length of stay" = c("2-9","10-19","20+"))))
art <- xtabs(~Treatment + Improved, data = Arthritis, subset = Sex == "Female")

set.seed(myseed)
art_max <- coindep_test(art, n = 5000)

set.seed(myseed)
ucb_max <- coindep_test(aperm(UCBAdmissions, c(3,2,1)), margin = "Dept", n = 5000)

@@

set.seed(myseed)
hos_chisq <- coindep_test(hospital, n = 5000, indepfun = function(x) sum(x^2))

@

@@

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \cite{vcd:SAS:2005}, \pkg{ViSta} \cite{vcd:young:1996}, 
\pkg{MANET} \cite{vcd:Unwin+Hawkins+Hofmann:1996},
\pkg{Mondrian} \cite{vcd:Theus:2003},
\proglang{R} \cite{vcd:R:2005}, and \proglang{S-PLUS} \cite{vcd:SPLUS:2005}. 
Implementations of association and sieve plots can only be found in
\proglang{R} and \proglang{SAS} (in the latter, these plots are available
for two-way tables only).
Table~\ref{tab:compare} gives an overview of the available functionality in these systems.
The figures in this chapter have all been produced using the
\proglang{R} system, using the extension packages \pkg{vcd} 
\cite{vcd:Meyer+Zeileis+Hornik:2005} and
\pkg{scatterplot3d} \cite{vcd:ligges+maechler:2003}
(Figure~\ref{fig:threedee} only), all freely available from the
Comprehensive \proglang{R} Archive Network
(\url{http://CRAN.R-project.org/}). The \proglang{R} code used for the
figures is available from \url{http://statmath.wu-wien.ac.at/projects/vcd/}.

@@

Although far from optimal, contingency tables are frequently visualized using
grouped bar plots (see Figure~\ref{fig:grouped}) 
or even by means of 3D-bar charts (see Figure~\ref{fig:threedee}). It seems hard
to detect the aforementioned pattern in these, especially in the 3D
plot where the perspective view tends to distort the true proportions
of the bars. In the following, we
will introduce three graphical methods that are better suited for
contingency tables.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<grouped,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
barplot(hospital, legend = rownames(hospital), beside = TRUE, 
        xlab = "Length of stay (years)", ylab = "Number of patients")
@
\caption{Bar plot for the \data{hospital} data.}
\label{fig:grouped}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<threedeebars,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
myhospital = t(hospital)[,3:1]
mydat = data.frame("Length of stay" = as.vector(row(myhospital)), 
                   "Visit frequency" = as.vector(col(myhospital)), 
                   "Number of patients" = as.vector(myhospital))
scatterplot3d(mydat, type = "h", pch = " ", lwd = 10, 
x.ticklabs = c("2-9","","10-19","","20+"), 
              y.ticklabs = c("Never","","Less than monthly","","Regular"), 
              xlab = "Length of stay (years)", ylab = "Visit frequency", zlab = "Number of patients",
              y.margin.add = 0.2,
              color = "black", box = F)
@
\caption{3D-bar chart for the \data{hospital} data.}
\label{fig:threedee}
\end{center}
\end{figure}

@@

Now, the expected cell frequencies in this model are simply
$\hat n_{ij} = n_{i+}n_{+j}/n$.
The expected table for our sample data is given in Table~\ref{tab:hospital2}.
It could again be visualized using a mosaic plot, this
time applied to the table of expected frequencies. 
If we cross-tabulate each tile to fill it with a number of squares equal to
the corresponding number of \emph{observed} frequencies, we get a \emph{sieve plot}
(see Figure~\ref{fig:sievefig}). It implicitly compares
expected and observed values since the density of the grid will
increase with the deviation of the observed from the expected
values. This allows the detection of general association patterns (for
nominal variables) and of linear association (for ordinal variables),
the latter producing tiles of either very high or very low density 
along one of the diagonals.
In the case of our data, the density of the rectangles is marked along the secondary
diagonal, indicating a negative association of 
the two variables. This gives evidence to the fact that for these patients,
visit frequency decreases with the length of stay.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<sievefig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
sieve(t(hospital), split = TRUE, pop = FALSE, gp = gpar(lty =
"dotted", col = "black"))
labeling_cells(text = t(hospital), clip = FALSE, gp = gpar(fontface = 2, fontsize = 15))(t(hospital))
@ 
\caption{Sieve plot for the \data{hospital} data.}
\label{fig:sievefig}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<colorfig,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,9), radius = 1, col = rainbow(9), labels = 360 * 0:8/9)
pie(rep(1,9), radius = 1, col = rainbow_hcl(9), labels = 360 * 0:8/9)
@ 
\caption{Qualitative color palette for the HSV (left) 
         and HCL (right) spaces. The HSV colors are $(H, 100, 100)$ and
	 the HCL colors $(H, 50, 70)$ for the same hues $H$. Note that
         in a monochrome version of this paper, all pies in the right
         wheel will be shaded with the same gray, i.e., appear identical.}
\label{fig:colors}
\end{center}
\end{figure}

@@

The heuristic for choosing the cut-off points 
in the Friendly shading may lead to wrong conclusions: especially in
large tables, the test of
independence may not be significant even though some of the residuals are
``large''. On the other hand, the test might be significant even
though the residuals are ``small''. In fact, the cut-off points are really
data-dependent. Consider the case of the \data{arthritis} data 
\cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, stratified by
gender (see Table~\ref{tab:arthritis} for the female patients):
\begin{table}[h]
\begin{center}
\begin{tabular}{|l|rrr|r|}
\hline
 & \multicolumn{4}{c|}{\textbf{Improved}}\\
\textbf{Treatment} & None & Some & Marked & $\Sigma$ \\
\hline
Placebo & 19 & 7 & 6 & 32 \\
Treatment & 6 & 5 & 16 & 27 \\\hline
$\Sigma$ & 25 & 12 & 22 & 59 \\
\hline
\end{tabular}
\caption{The \data{arthritis} data (female patients).}
\label{tab:arthritis}
\end{center}
\end{table}
Figure~\ref{fig:arthritis} visualizes the results for the female
patients, again by means of a mosaic display. Clearly, the hypothesis
of independence is rejected by the $\chi^2$ test even on a $1\%$ level 
($p=\Sexpr{round(art_chisq$p.value,4)}$), but
since all residuals are in the 
$[\Sexpr{round(min(chisq.test(art)$residuals),4)},\Sexpr{round(max(chisq.test(art)$residuals),4)}]$ %$
interval, the tiles remain uncolored. A solution to this issue is to use a different
test statistic, for example the maximum of the absolute values of the
Pearson residuals \cite{vcd:Meyer+Zeileis+Hornik:2003} instead of the sum of squares:

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<Arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(art, shade = TRUE, gp_args = list(p.value = art_chisq$p.value))
@
\caption{Mosaic plot for the \data{arthritis} data, using the $\chi^2$ test 
and fixed cut-off points for the shading.}
\label{fig:arthritis}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<Arthritis2fig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
set.seed(myseed)
mosaic(art, gp = shading_max, gp_args = list(n = 5000))
@
\caption{Mosaic plot for the \data{arthritis} data, using the maximum test and 
data-driven cut-off points for the residuals.}
\label{fig:arthritis2}
\end{center}
\end{figure}

@@

As an example, consider the well-known UCB admissions data
\cite{vcd:Bickel+Hammel+O'Connell:1975} on applicants, classified by
admission and gender, to graduate school at the University of California
in Berkeley for the six largest departments in 1973. The `flattened'
contingency table, putting department into the columns and
admit---nested in gender---into the rows, is given by
Table~\ref{tab:ucb}. The table aggregated over all departments gives
wrong evidence of gender bias, i.e., higher admission rates for male
students.

\begin{table}[ht]
\begin{center}
\begin{tabular}{|ll|rrrrrr|}
\hline
 & & \multicolumn{6}{c|}{\textbf{Department}} \\
\textbf{Gender} & \textbf{Admit} & A & B & C & D & E & F\\\hline
Male   & Admitted & 512 & 353 & 120 & 138 & 53 & 22\\
       & Rejected & 313 & 207 & 205 & 279 & 138 & 351\\
Female & Admitted & 89 & 17 & 202 & 131 & 94 & 24 \\
       & Rejected & 19 & 8 & 391 & 244 & 299 & 317 \\
\hline
\end{tabular}
\caption{The \data{UCB admissions} data, in flat representation.}
\label{tab:ucb}
\end{center}
\end{table}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<doubleucb,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
doubledecker(aperm(aperm(UCBAdmissions, c(1,3,2))[2:1,,], c(2,3,1)), margins = c(left = 0, right = 5),col = rev(grey.colors(2)))
@
\caption{Doubledecker plot for the \data{UCBAdmissions} data.}
\label{fig:doubleucb}
\end{center}
\end{figure}

@@

Another approach is to visualize (the deviations from) a particular fitted model 
with similar ideas as in Section~\ref{sec:extensions}. Thus, mosaic displays can
be extended from purely explorative views to model-based views on the data by
residual-based shadings and association plots can directly visualize deviations
from a given model. Typically, the models considered are log-linear models, however
residual-based shadings are also conceivable for other types of models such as logistic
regression.
In our example, from the exploratory analysis, we already suspect that
admission and gender are independent, given department, except for
department A. 
Therefore, a sensible way of representing the table is to 
nest admission into department instead of gender to obtain a stratified view.
The corresponding mosaic and association plots are shown
in Figures~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}. The shading visualizes the 
model of admission and gender being conditionally independent, given department. Clearly,
there is no gender bias in the departments except for department A: here, 
significantly \textit{more} female students are admitted then expected under independence.
Mosaic displays are already recognized as an excellent
visualization tool for log-linear models 
\cite{vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
Using the flat table representation, the association plot similarly extends from
the simple two-way case to a structured residual plot usable for 
the diagnostics of log-linear models. Especially if a
mosaic display contains very small or empty cells, the corresponding association plot
might provide an easier way to detect deviation patterns than the
residual-based shading added to the mosaic plot. Yet another approach
is to plot a mosaic plot of the \emph{expected} frequencies and to highlight
the (relative) residuals in the cells, using different colors for
positive and negative residuals \cite{vcd:Theus+Lauer:1999}.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<ucbmosaic,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
mosaic(aperm(UCBAdmissions, c(3,2,1)), data = UCBAdmissions, split = TRUE,
shade = TRUE, keep = FALSE, residuals = ucb_max$residuals,
gp_args = list(p.value = ucb_max$p.value), rep = c(Admit = FALSE)) 
@
\caption{Mosaic plot for the \data{UCB admissions} data.}
\label{fig:ucbmosaic}
\end{center}
\end{figure}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<ucbassoc,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
assoc(aperm(UCBAdmissions, c(3,2,1)), split = c(TRUE,FALSE,TRUE), shade = TRUE, 
residuals_type = "Pearson", residuals = ucb_max$residuals,
gp_args = list(p.value = ucb_max$p.value), rep = c(Admit = FALSE))
@
\caption{Association plot for the \data{UCB admissions} data.}
\label{fig:ucbassoc}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
cotabplot(aperm(UCBAdmissions, c(2,1,3)), panel = cotab_coindep, shade = TRUE, 
          legend = FALSE, 
          panel_args = list(type = "assoc", margins = c(2,1,1,2), varnames = FALSE))
@
\caption{Conditional association plot for the \data{UCB admissions}
  data. For each individual association plot, gender is in the rows
  and admit is in the columns.}
\label{fig:ucbadmissions}
\end{center}
\end{figure}

@@

%FIXME% Better structure:
%FIXME%   o dependent variable: attitude
%FIXME%   o dependence on age and education is rather clear
%FIXME%     (approval of moderate punishment increases with age and
%FIXME%     decreases with education)
%FIXME%   o question is conditional independence of attitude and
%FIXME%     memory given the other margins
%FIXME%   o hence for exploratory analysis: split in conditioning
%FIXME%     variables (age, education) first, then explanatory variable (memory),
%FIXME%     then dependent variable (attitude) for which highlighting is used.
%FIXME%   o for model-based view: suppress marginal distribution of
%FIXME%     conditioning variables (and their interaction), split first
%FIXME%     in explanatory variable in each cell, use shading based on
%FIXME%     residuals in conditional independence model.
%FIXME%   o advantage of exploratory view: joint distribution of all variables
%FIXME%     can be seen. disadvantage: visualization strongly influence
%FIXME%     by the marginal distribution of education over age (in particular
%FIXME%     the large proportion of 40+ people with elementary education)
%FIXME%     which is not of interest in the conditional independence problem.

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<punishhigh,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
tab <- xtabs(Freq ~ memory + attitude + age + education, data = Punishment)
dimnames(tab)[[3]][3] <- "40+"
mosaic(~ age + education + memory + attitude, data = tab, keep = FALSE,
       gp = gpar(fill = grey.colors(2)), spacing = spacing_highlighting, 
       rep = c(attitude = FALSE))
@
\caption{Mosaic plot with highlighting for the \data{punishment} data.}
\label{fig:punishhigh}
\end{center}
\end{figure}

@@

% ## \setkeys{Gin}{width=0.7\textwidth}
% ## \begin{figure}
% ## \begin{center}
% ## <<punishmosaic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
% ## tab_test <- coindep_test(tab, margin = c(3, 4))
% ## mosaic(~ age + education + memory + attitude, data = tab, keep = FALSE,
% ##        expected = ~ (memory + attitude) * education * age,
% ##        gp_args = list(interpolate = tab_test$qdist(c(0.9, 0.99))), 
% ##        rep = c(attitude = F), set_varnames = c(attitude = "attitude to punishment"),
% ##        set_labels = list(age = c("15-24","25-39","40+")))
% ## 
% ## \caption{Mosaic plot for the \data{punishment} data.}
% ## \label{fig:punishmosaic}
% ## \end{center}
% ## \end{figure}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<punishcotab,fig=TRUE,echo=FALSE,height=10,width=10,results=hide>>=
set.seed(123)
cotabplot(tab, panel = cotab_coindep, panel_args = list(varnames = FALSE, margins = c(2, 1, 1, 2)))
@
\caption{Conditional mosaic plot for the \data{punishment} data. For
  each mosaic plot, memory is in the rows (first split) and attitude
  in the columns (second split).}
\label{fig:punishcotab}
\end{center}
\end{figure}

>>>>

<<preliminaries,echo=FALSE,results=hide>>=
## random seed
myseed <- 1071
set.seed(myseed)

## packages
library("vcd")
library("scatterplot3d")
library("xtable")

## data
hospital <- as.table(matrix(c(43,16,3,6,11,10,9,18,16), c(3,3), byrow = TRUE, 
  dimnames = list("Visit frequency" = c("Regular","Less than monthly","Never"), 
                  "Length of stay" = c("2-9","10-19","20+"))))

data("Arthritis")
art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female")
names(dimnames(art))[2] <- "Improvement"

data("UCBAdmissions")
names(dimnames(UCBAdmissions)) <- c("Admission", "Gender", "Department")

data("Punishment")
punish <- xtabs(Freq ~ memory + attitude + age + education, data = Punishment)
dimnames(punish)[[3]][3] <- "40+"

## inference
set.seed(myseed)
art_max <- coindep_test(art, n = 5000)

set.seed(myseed)
ucb_max <- coindep_test(aperm(UCBAdmissions, c(3,2,1)), margin = "Department", n = 5000)

@@

set.seed(myseed)
hos_chisq <- coindep_test(hospital, n = 5000, indepfun = function(x) sum(x^2))
@

@@

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \cite{vcd:SAS:2005}, \pkg{ViSta} \cite{vcd:young:1996}, 
\pkg{MANET} \cite{vcd:Unwin+Hawkins+Hofmann:1996},
\pkg{Mondrian} \cite{vcd:Theus:2003},
\proglang{R} \cite{vcd:R:2005}, and \proglang{S-PLUS} \cite{vcd:SPLUS:2005}. 
Implementations of association and sieve plots can only be found in
\proglang{R} and \proglang{SAS} (in the latter, these plots are available
for two-way tables only).
Table~\ref{tab:compare} gives an overview of the available functionality in these systems.
The figures in this chapter have all been produced using the
\proglang{R} system, using the extension packages \pkg{vcd} 
\cite{vcd:Meyer+Zeileis+Hornik:2005} and
\pkg{scatterplot3d} \cite{vcd:ligges+maechler:2003}
(Figure~\ref{fig:bars3d} only), all freely available from the
Comprehensive \proglang{R} Archive Network
(\url{http://CRAN.R-project.org/}). The \proglang{R} code used for the
figures is available from \url{http://statmath.wu-wien.ac.at/projects/vcd/}.

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<barsgrouped,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
barplot(hospital, legend = rownames(hospital), beside = TRUE, 
        xlab = "Length of stay", ylab = "Number of patients")
@
\caption{Bar plot for the \data{hospital} data.}
\label{fig:barsgrouped}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<bars3d,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
myhospital = t(hospital)[,3:1]
mydat = data.frame("Length of stay" = as.vector(row(myhospital)), 
                   "Visit frequency" = as.vector(col(myhospital)), 
                   "Number of patients" = as.vector(myhospital))
scatterplot3d(mydat, type = "h", pch = " ", lwd = 10, 
x.ticklabs = c("2-9","","10-19","","20+"), 
              y.ticklabs = c("Never","","Less than monthly","","Regular"), 
              xlab = "Length of stay", ylab = "Visit frequency", zlab = "Number of patients",
              y.margin.add = 0.2,
              color = "black", box = F)
@
\caption{3D-bar chart for the \data{hospital} data.}
\label{fig:bars3d}
\end{center}
\end{figure}

Although far from optimal, contingency tables are frequently visualized using
grouped bar plots (see Figure~\ref{fig:barsgrouped}) 
or even by means of 3D-bar charts (see Figure~\ref{fig:bars3d}). It seems hard
to detect the aforementioned pattern in these, especially in the 3D
plot where the perspective view tends to distort the true proportions
of the bars. In the following, we
will introduce three graphical methods that are better suited for
contingency tables.

@@

Now, the expected cell frequencies in this model are simply
$\hat n_{ij} = n_{i+}n_{+j}/n$.
The expected table for our sample data is given in Table~\ref{tab:hospital2}.
It could again be visualized using a mosaic plot, this
time applied to the table of expected frequencies. 
If we cross-tabulate each tile to fill it with a number of squares equal to
the corresponding number of \emph{observed} frequencies, we get a \emph{sieve plot}
(see Figure~\ref{fig:sieve}). It implicitly compares
expected and observed values since the density of the grid will
increase with the deviation of the observed from the expected
values. This allows the detection of general association patterns (for
nominal variables) and of linear association (for ordinal variables),
the latter producing tiles of either very high or very low density 
along one of the diagonals.
In the case of our data, the density of the rectangles is marked along the secondary
diagonal, indicating a negative association of 
the two variables. This gives evidence to the fact that for these patients,
visit frequency decreases with the length of stay.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<sieve,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
sieve(t(hospital), split = TRUE, pop = FALSE, gp = gpar(lty =
"dotted", col = "black"))
labeling_cells(text = t(hospital), clip = FALSE, gp = gpar(fontface = 2, fontsize = 15))(t(hospital))
@ 
\caption{Sieve plot for the \data{hospital} data.}
\label{fig:sieve}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<color,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,9), radius = 1, col = rainbow(9), labels = 360 * 0:8/9)
pie(rep(1,9), radius = 1, col = rainbow_hcl(9), labels = 360 * 0:8/9)
@ 
\caption{Qualitative color palette for the HSV (left) 
         and HCL (right) spaces. The HSV colors are $(H, 100, 100)$ and
	 the HCL colors $(H, 50, 70)$ for the same hues $H$. Note that
         in a monochrome version of this paper, all pies in the right
         wheel will be shaded with the same gray, i.e., appear identical.}
\label{fig:colors}
\end{center}
\end{figure}

@@

The heuristic for choosing the cut-off points 
in the Friendly shading may lead to wrong conclusions: especially in
large tables, the test of
independence may not be significant even though some of the residuals are
``large''. On the other hand, the test might be significant even
though the residuals are ``small''. In fact, the cut-off points are really
data-dependent. Consider the case of the \data{arthritis} data 
\cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, stratified by
gender (see Table~\ref{tab:arthritis} for the female patients):
\begin{table}[h]
\begin{center}
\begin{tabular}{|l|rrr|r|}
\hline
 & \multicolumn{4}{c|}{\textbf{Improvement}}\\
\textbf{Treatment} & None & Some & Marked & $\Sigma$ \\
\hline
Placebo & 19 & 7 & 6 & 32 \\
Treatment & 6 & 5 & 16 & 27 \\\hline
$\Sigma$ & 25 & 12 & 22 & 59 \\
\hline
\end{tabular}
\caption{The \data{arthritis} data (female patients).}
\label{tab:arthritis}
\end{center}
\end{table}
Figure~\ref{fig:arthritis} visualizes the results for the female
patients, again by means of a mosaic display. Clearly, the hypothesis
of independence is rejected by the $\chi^2$ test even on a $1\%$ level 
($p=\Sexpr{round(art_chisq$p.value,4)}$), but
since all residuals are in the 
$[\Sexpr{round(min(chisq.test(art)$residuals),4)},\Sexpr{round(max(chisq.test(art)$residuals),4)}]$ %$
interval, the tiles remain uncolored. A solution to this issue is to use a different
test statistic, for example the maximum of the absolute values of the
Pearson residuals \cite{vcd:Meyer+Zeileis+Hornik:2003} instead of the sum of squares:

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(art, shade = TRUE, gp_args = list(p.value = art_chisq$p.value))
@
\caption{Mosaic plot for the \data{arthritis} data, using the $\chi^2$ test 
and fixed cut-off points for the shading.}
\label{fig:arthritis}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<arthritis2,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
set.seed(myseed)
mosaic(art, gp = shading_max, gp_args = list(n = 5000))
@
\caption{Mosaic plot for the \data{arthritis} data, using the maximum test and 
data-driven cut-off points for the residuals.}
\label{fig:arthritis2}
\end{center}
\end{figure}

@@

As an example, consider the well-known UCB admissions data
\cite{vcd:Bickel+Hammel+O'Connell:1975} on applicants, classified by
admission and gender, to graduate school at the University of California
in Berkeley for the six largest departments in 1973. The `flattened'
contingency table, putting department into the columns and
admission---nested in gender---into the rows, is given by
Table~\ref{tab:ucb}. The table aggregated over all departments gives
wrong evidence of gender bias, i.e., higher admission rates for male
students.

\begin{table}[ht]
\begin{center}
\begin{tabular}{|ll|rrrrrr|}
\hline
 & & \multicolumn{6}{c|}{\textbf{Department}} \\
\textbf{Gender} & \textbf{Admission} & A & B & C & D & E & F\\\hline
Male   & Admitted & 512 & 353 & 120 & 138 & 53 & 22\\
       & Rejected & 313 & 207 & 205 & 279 & 138 & 351\\
Female & Admitted & 89 & 17 & 202 & 131 & 94 & 24 \\
       & Rejected & 19 & 8 & 391 & 244 & 299 & 317 \\
\hline
\end{tabular}
\caption{The \data{UCB admissions} data, in flat representation.}
\label{tab:ucb}
\end{center}
\end{table}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<ucbhigh,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
doubledecker(aperm(aperm(UCBAdmissions, c(1,3,2))[2:1,,], c(2,3,1)), margins = c(left = 0, right = 5),col = rev(grey.colors(2)))
@
\caption{Doubledecker plot for the \data{UCBAdmissions} data.}
\label{fig:doubleucb}
\end{center}
\end{figure}

@@

Another approach is to visualize (the deviations from) a particular fitted model 
with similar ideas as in Section~\ref{sec:extensions}. Thus, mosaic displays can
be extended from purely explorative views to model-based views on the data by
residual-based shadings and association plots can directly visualize deviations
from a given model. Typically, the models considered are log-linear models, however
residual-based shadings are also conceivable for other types of models such as logistic
regression.
In our example, from the exploratory analysis, we already suspect that
admission and gender are independent, given department, except for
department A. 
Therefore, a sensible way of representing the table is to 
nest admission into department instead of gender to obtain a stratified view.
The corresponding mosaic and association plots are shown
in Figures~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}. The shading visualizes the 
model of admission and gender being conditionally independent, given department.
\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<ucbmosaic,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
mosaic(aperm(UCBAdmissions, c(3,2,1)), data = UCBAdmissions, split = TRUE,
shade = TRUE, keep = FALSE, residuals = ucb_max$residuals,
gp_args = list(p.value = ucb_max$p.value), rep = c(Admission = FALSE)) 
@
\caption{Mosaic plot for the \data{UCB admissions} data.}
\label{fig:ucbmosaic}
\end{center}
\end{figure}
\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<ucbassoc,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
assoc(aperm(UCBAdmissions, c(3,2,1)), split = c(TRUE,FALSE,TRUE), shade = TRUE, 
residuals_type = "Pearson", residuals = ucb_max$residuals,
gp_args = list(p.value = ucb_max$p.value), rep = c(Admission = FALSE))
@
\caption{Association plot for the \data{UCB admissions} data.}
\label{fig:ucbassoc}
\end{center}
\end{figure}
Clearly, there is no gender bias in the departments except for department A: here, 
significantly \textit{more} female students are admitted then expected under independence.
Mosaic displays are already recognized as an excellent
visualization tool for log-linear models 
\cite{vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
Using the flat table representation, the association plot similarly extends from
the simple two-way case to a structured residual plot usable for 
the diagnostics of log-linear models. Especially if a
mosaic display contains very small or empty cells, the corresponding association plot
might provide an easier way to detect deviation patterns than the
residual-based shading added to the mosaic plot. Yet another approach
is to plot a mosaic plot of the \emph{expected} frequencies and to highlight
the (relative) residuals in the cells, using different colors for
positive and negative residuals \cite{vcd:Theus+Lauer:1999}.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<ucbcotab,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
cotabplot(aperm(UCBAdmissions, c(2,1,3)), panel = cotab_coindep, shade = TRUE, 
          legend = FALSE, 
          panel_args = list(type = "assoc", margins = c(2,1,1,2), varnames = FALSE))
@
\caption{Conditional association plot for the \data{UCB admissions}
  data. For each individual association plot, gender is in the rows
  and admission is in the columns.}
\label{fig:ucbadmissions}
\end{center}
\end{figure}

@@

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<punishhigh,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(~ age + education + memory + attitude, data = punish, keep = FALSE,
       gp = gpar(fill = grey.colors(2)), spacing = spacing_highlighting, 
       rep = c(attitude = FALSE))
@
\caption{Mosaic plot with highlighting for the \data{punishment} data.}
\label{fig:punishhigh}
\end{center}
\end{figure}

@@

% ## \setkeys{Gin}{width=0.7\textwidth}
% ## \begin{figure}
% ## \begin{center}
% ## <<punishmosaic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
% ## punish_test <- coindep_test(punish, margin = c(3, 4))
% ## mosaic(~ age + education + memory + attitude, data = punish, keep = FALSE,
% ##        expected = ~ (memory + attitude) * education * age,
% ##        gp_args = list(interpolate = punish_test$qdist(c(0.9, 0.99))), 
% ##        rep = c(attitude = F), set_varnames = c(attitude = "attitude to punishment"),
% ##        set_labels = list(age = c("15-24","25-39","40+")))
% ## 
% ## \caption{Mosaic plot for the \data{punishment} data.}
% ## \label{fig:punishmosaic}
% ## \end{center}
% ## \end{figure}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<punishcotab,fig=TRUE,echo=FALSE,height=10,width=10,results=hide>>=
set.seed(123)
cotabplot(punish, panel = cotab_coindep, panel_args = list(varnames = FALSE, margins = c(2, 1, 1, 2)))
@
\caption{Conditional mosaic plot for the \data{punishment} data. For
  each mosaic plot, memory is in the rows (first split) and attitude
  in the columns (second split).}
\label{fig:punishcotab}
\end{center}
\end{figure}
